
\title{Measuring What Matters: The AI Pluralism Index}

\author{
Rashid Mushkani\\
Université de Montréal, Mila – Quebec AI Institute\\
\texttt{rashid.ahmad.mushkani@umontreal.ca}
}

\begin{document}

\maketitle

\begin{abstract}
AI systems increasingly shape knowledge, communication, and decisions, while development is concentrated in a few firms and states. Capability benchmarks abound, yet public measures of pluralistic governance are scarce. We define \emph{AI pluralism} as conditions under which affected stakeholders influence system objectives, data practices, guardrails, and deployment. We introduce the AI Pluralism Index (AIPI), an evidence-based framework that evaluates producers and system families along four pillars: participatory governance, inclusivity and diversity, transparency, and accountability. AIPI aggregates observable practices from verifiable artifacts and external evaluations, with an open, versioned protocol for community proposals and adjudication. We outline the measurement model, coding scheme, and data pipeline spanning structured web search, repository analysis, and expert interviews, and we describe validation via inter-rater reliability and sensitivity analysis. We illustrate with descriptive statistics from two public-interest AI corpora and situate AIPI relative to adjacent transparency and safety indices. The index aims to steer incentives toward pluralistic practice and provide comparable evidence for policymakers, procurers, and the public.
\end{abstract}

\end{document}
